---
output:
  md_document:
    variant: markdown_github
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

## Overview

sppt is an [R](https://www.r-project.org) package that implements Martin 
Andresen's Spatial Point Pattern Test. It has been published [here](http://jrc.sagepub.com/content/48/1/58) (and elsewhere). The test 
measures the degree of similarity at the local level between two spatial point 
patterns and is an area-based test.

This spatial point pattern test is not for the purpose of testing point 
patterns with the null hypotheses of random, uniform, or clustered 
distributions, but may be used to compare a particular point pattern with 
these distributions. One advantage of the test is that it can be performed for 
a number of different area boundaries using the same original point datasets.

In February 2018 a new test was added to the package courtesy of Andrew Wheeler,
as a result of the following publication:

> Wheeler, A., Steenbeek, W., & Andresen, M.A. (2018). Testing for Similarity in Area-Based Spatial Patterns: Alternative Methods to Andresen's Spatial Point Pattern Test. (SSRN: https://papers.ssrn.com/sol3/papers.cfm?abstract_id=3111822)

Therefore the most recent version of the package has changed from version 0.1.4
to version 0.1.5. For version 0.1.4, please cite package 'sppt' in publications as:

> Steenbeek, W., Vandeviver, C. Andresen, M.A., Malleson, N. (2017). *sppt: Spatial Point Pattern Test*. R package version 0.1.4. URL: https://github.com/wsteenbeek/sppt

For version 0.1.5, the updated reference is:

> Steenbeek, W., Vandeviver, C. Andresen, M.A., Malleson, N., Wheeler, A. (2018). *sppt: Spatial Point Pattern Test*. R package version 0.1.5. URL: https://github.com/wsteenbeek/sppt

## Installation

You can install the package from this 
[GitHub repository](https://github.com/wsteenbeek/sppt). You first need to
install the [devtools](https://CRAN.R-project.org/package=devtools) package.

```{r, eval = FALSE}
install.packages("devtools")
```

Then install sppt using the `install_github` function in the
[devtools](https://CRAN.R-project.org/package=devtools) package.

```{r, eval = FALSE}
devtools::install_github("wsteenbeek/sppt")
```

## Example

Spatial objects areas.sp, points1.sp, and points2.sp, are included in the 
package. For example, you can think of these as neighborhoods and the locations
where crimes occur in two different years.

```{r, eval = TRUE, warning = FALSE, message = FALSE}
library(sppt)
plot(areas.sp)
points(points1.sp, col="blue", pch = 19)
points(points2.sp, col="red", pch = 15)
```

The original function within the sppt package is also called `sppt`:

```{r, eval = TRUE}
set.seed(93255) # set seed for reproducibility
output <- sppt(points1.sp, points2.sp, areas.sp)
```

Two other functions were added in February 2018. `sppt_boot()`:

```{r, eval = TRUE}
set.seed(93255) # set seed for reproducibility
output2 <- sppt_boot(points1.sp, points2.sp, areas.sp)
```

and `sppt_diff()`:

```{r, eval = TRUE}
set.seed(93255) # set seed for reproducibility
output3 <- sppt_diff(points1.sp, points2.sp, areas.sp)
```

You can see the results of the test by inspecting the SpatialPolygonsDataFrame, for example:

```{r, eval = TRUE}
output@data
```

## Vignettes

The package includes vignettes explaining the sppt procedure in more detail;
a worked example of a toy dataset and actual crime data; a comparison
between this R package and an existing Java application that has been written by 
[Nick Malleson](http://nickmalleson.co.uk/); and the new functions `sppt_boot()`
and `sppt_diff()`.

You can view which vignettes are available using:

```{r, eval = FALSE}
browseVignettes("sppt")
```

To save time and because they may require additional packages, `devtools::install_github()` will not build vignettes by default. If the command above returns that vignettes are not found, you need to take a few additional steps.

1. Install the rmarkdown package with `install.packages("rmarkdown")`

2. [Install pandoc](http://johnmacfarlane.net/pandoc/installing.html) (and afterwards restart your computer)

3. Then, install the package again but force building of the vignettes using `devtools::install_github("wsteenbeek/sppt", build_vignettes = TRUE, force = TRUE)`. This will take a few minutes.

You should now be able to view which vignettes are available using:

```{r, eval = FALSE}
browseVignettes("sppt")
```

To directly read the vignettes rather than going through `browseVignettes("sppt")` you can use:

```{r, eval = FALSE}
vignette("sppt", package = "sppt")
vignette("sppt_comparison", package = "sppt")
vignette("sppt_diff", package = "sppt")
```

If the above still doesn't work, then a quick-fix is to just view the vignettes online using the [GitHub HTML Preview service](http://htmlpreview.github.io/):

1. [sppt](http://htmlpreview.github.io/?https://github.com/wsteenbeek/sppt/blob/master/inst/doc/sppt.html)

2. [sppt_comparison](http://htmlpreview.github.io/?https://github.com/wsteenbeek/sppt/blob/master/inst/doc/sppt_comparison.html)

3. [sppt_diff](http://htmlpreview.github.io/?https://github.com/wsteenbeek/sppt/blob/master/inst/doc/sppt_diff.html)


## License

This package is free software; you can redistribute it and/or modify it
under the terms of the GNU General Public License, version 3, as
published by the Free Software Foundation.

This program is distributed in the hope that it will be useful, but
without any warranty; without even the implied warranty of
merchantability or fitness for a particular purpose.  See the GNU
General Public License for more details.

A copy of the GNU General Public License, version 3, is available at
<https://www.r-project.org/Licenses/GPL-3>
